\chapter*{Acknowledgement}

I would like to express my sincere gratitude and appreciation to all those who have supported and contributed to the completion of my master's thesis. Without their invaluable assistance, encouragement, and guidance, this research would not have been possible.

First and foremost, I am deeply indebted to my thesis advisors, Dr. Alan Edelman and Dr. Chris Rackauckas, for their unwavering support, patience, and expertise throughout this journey. Their insightful feedback, constructive criticism, and constant encouragement played a pivotal role in shaping the direction and quality of this thesis. I am truly grateful for their mentorship and the invaluable lessons I have learned under their guidance.

I am also thankful to my peers at the Julia Lab CSAIL and the broader EECS MIT community, who provided a nurturing academic environment that facilitated my research endeavors. Their commitment to excellence and dedication to fostering intellectual growth have been instrumental in shaping my academic development. I extend my heartfelt appreciation to all my professors, colleagues, and peers who engaged in thought-provoking discussions, provided valuable feedback, and offered their support.

My gratitude extends to my parents, Abhijit Pal and Rupa Pal, for their unwavering love, belief in my abilities, and constant encouragement. They have been my pillars of strength throughout this journey. I am especially thankful to my brother, Dr. Aritra Pal, whose motivation to pursue a Ph.D. program paved the way for my academic journey. Without their constant support, understanding, and encouragement, this work would not have been possible.

Completing this master's thesis has been a challenging yet rewarding journey, and I feel truly honored to have had the opportunity to undertake this research. The support and contributions of everyone mentioned above have been pivotal in making this thesis a reality. I express my deepest gratitude to all for their unwavering support, belief in my abilities, and for being an integral part of this significant milestone in my academic and personal life.